Co-evolution of lexical meaning \& pragmatic use

### 2018-03-10 paper/cogsci-submission2.tex
minor polishing of first parts; added paragraph on co-evolution

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses, at least since \citeposs{zipf:1949} explanation of word
frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another, as natural languages need to be learnable to survive their
faithful transmission across generations. Furthermore, the effects of even small learning biases implicit in acquisition can build up and have quite striking impact on an evolving language in a process of iterated
learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

@@

\subsection{The replicator mutator dynamic}
An elegant formal approach to capture the interaction between communicative efficiency and
learnability is the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. The
RMD is a population dynamic. It describes the average change in the composition of a community
of agents of different types. The dynamic assumes a virtually infinite population, so that it
is able to abstract away from what individual agents do and how they change their individual
behavior over time. The result is an abstract description of the most probable trajectory of
change in the population vector $\vec{x}$, which codifies relative frequencies of types, so
that $x_i$ is the relative frequency of type $i$ and $\sum\nolimits_i x_i = 1$. In its
simplest discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$ in
the population at the next time step as a function of: (i) the frequency $x_i$ of each type $i$
before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and (iii)
the probability $Q_{ji}$ that an agent who observes the behavior of an agent with type $j$ ends
up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j \frac{x_jf_j Q_{ji}}{\Phi}\,,
\end{align}
where $\Phi = \sum_k x_k f_k$ is the average fitness in the population.

The RMD consists of two components: fitness-based selection and transmission biases encoded in
the so-called mutation matrix $Q$. This becomes most transparent when we consider an equivalent
formulation in terms of a step-wise application of the discrete-time replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$ and its
subsequent multiplication with the mutation matrix $Q$:
\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\Phi}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
%
If there is no mutation, so that types are always replicated faithfully with $Q_{ji}=1$
whenever $j=i$, the dynamic reduces to the replicator dynamic. If fitness is the same for all
types, $f_i = f_j$ for all $i$ and $j$, the dynamics reduces to a process of iterated
mutation. Appendix~\ref{sec:minim-example-repl} provides an abstract minimal example of how the
replicator mutator dynamic works.

@@

According to the replicator dynamic the relative frequency $x_i$ of type $i$ increases
proportional to its average fitness in the population. This dynamic is popular and versatile
because it can be derived from many abstract processes of biological and cultural transmission
and selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}. Under a biological interpretation, the model describes
population-level changes when the number of offspring under single-parent preproduction is
proportional to the parent's fitness. Under a cultural interpretation, examples of concrete
agent-level processes that lead to population-level changes as described by the replicator
dynamic include conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. In
conditional imitation, for instance, each agent carries a fixed type $i$ which it may change to
a type $j$ when it happens to observe another agent using type $j$. Imitation occurs with a
probability that is a monotonic function of how well $j$ fares against the population: the
better the behavior $j$, the more likely it is to be imitated.\footnote{A crucial assumption in
  all of the derivations of the replicator dynamic as a biological or cultural process of
  evolution is that agents themselves are rather limited in their strategic capacities. They
  are, for example, not able to innovate a strategy, i.e., to just choose a type, that is
  optimal given the current population state. They are also not able to strategically adapt
  their behavior to any particular agent they are interacting with on some given occasion. Such
  minimalism about the strategic capacities of agents is a conservative methodological
  choice. It allows us to study ---as the formal thought experiment that any such model is---
  how fitness-based selection changes a population in the absence of optimal agent-level
  reasoning and it abstracts away from specific assumptions about how fitness-based selection
  might be realized concretely.}

@@

In the following application to language evolution, we prefer a cultural interpretation of the
mutator component as the perturbation induced by how easy different linguistic traits are
learnable from observation of language use. Nothing of current relevance hinges on whether the
replicator component is interpreted as a biological process or a cultural process. We believe
that this flexibility of interpretation is an asset. It abstracts away from specific
assumptions and focuses on a general, albeit abstract description of fitness-based selection in
general. Still, to have something concrete in mind when interpreting the following application,
it is also important to be concrete. Appendix~\ref{sec:deriv-repl-mutat} therefore derives the
RMD from a specific scheme of agent-level conditional imitation.

@@

Under the replicator dynamic the proportion of type $i$ in a population will increase or
decrease as a function of its relative fitness $f_i$. In the context of language evolution,
fitness is usually associated with the ability to successfully communicate with other language
users from the same population \citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000,
  nowak+etal:2002}. Concretely, the fitness of type $i$ is its average expected communicative
success, or \emph{expected utility} (EU), given the relative frequencies $x_j$ of types $j$ in
the current population:
\begin{align*}
  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
\end{align*}
In signaling games, the expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
average success of $i$ when talking or listening to $j$. If, as standardly assumed, agents are speakers half of the
time, this yields:
\begin{align*}
  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, \text{EU}_S(t_i,t_j) + \nicefrac{1}{2} \, \text{EU}_H(t_i,t_j)\,,
\end{align*}
where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
speaker and as a hearer when communicating with $j$, defined as follows, where $n_i$ and $n_j$
are type $i$'s and type $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
\begin{flalign*}
  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} H_{n_j}(s' \mid m;L_j)
  \delta(s,s') \,, \\
 & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)\,.
\end{flalign*}
As usual, we assume that agents are cooperative: $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.

@@

Communicative success and learnability are central to the cultural evolution of language. These
components can be modelled, respectively, as replication based on a measure of fitness in terms
of communicative efficiency and iterated Bayesian learning. Their interaction is described by
the discrete time replicator mutator dynamic in (\ref{eq:RMD_discrete}), repeated here:
\begin{align*}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_h x_h f_h}\,.
\end{align*}
This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
frequency $x_i$ before the step, its fitness $f_i$, and the probability that a learner infers
$i$ when observing the behavior of a type-$j$ agent. Fitness-based selection is here thought of
not as biological (fitness as expected relative number of offspring) but cultural (fitness as
likelihood of being imitated or repeated) evolution, since the types that the dynamic operates
on are pairs consisting of a lexicon and a pragmatic use pattern. A type's communicative success depends
on how well it communicates within a population while its learnability depends on the
fidelity by which it is inferred by new generations of learners. The learners' task is
consequently to perform a joint inference over types of linguistic behavior and lexical
meaning.

@@

The model of the previous section formalizes the evolutionary competition between different
sets of lexical representations and ways of using them. This section looks at a case study on
scalar implicatures. It engages in a formal thought experiment to address the
question: if a population of language users could freely combine different lexica with
different dispositions to act on them, what are conditions under which the majority view of scalar
implicatures could have evolved?

Recall that the majority view is that scalar implicatures are non-lexicalized pragmatic
enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may} are
semantically weak expressions for which logically stronger expressions are salient, e.g.,
\emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em all}. If
the sentence ``Chris owns all of Johnny Cash's albums'' is true, then ``Chris owns some of
Johnny Cash's albums'' is also true. However, while weaker expressions such as {\em some} are
truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what
their use is normally taken to convey. Instead, the use of a less informative expression when a
more informative one could have been used can license a defeasible inference that stronger
alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). In this way, ``Chris owns some
of Johnny Cash's albums'' is strengthened to convey that she owns {\em some but not all}
albums. According to the majority view, this is a pragmatic inference, not part of the
conventional meaning.

In the following we consider a specific application of the model from Section~\ref{sec:model}
which allows us to address the question if or when scalar inferences might (not) lexicalize. We
consider what is perhaps one of the simplest non-trivial setups that speak to this matter and
reflect on their limitations in Section~\ref{sec:discussion}. Section
\ref{sec:setup:-stat-conc} first considers a minimal example intended to isolate effects of
replication and mutation and to gain insights into the sense in which semantic meaning and
pragmatic use can be said to co-evolve. We then turn to a larger and more realistic setup in
Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes simulations
for this case and their results.

\section{Co-evolution of semantics \& pragmatics in a restricted type space}
\label{subsec:example}

@@

We focus only on the four types from Section~\ref{sec:languages+use}. Recall that these types
are made up of a combination of either lexicon $\Llack$ or $\Lbound$ and either literal
(level-$0$) or pragmatic (level-$1$) behavior.

@@

From Section~\ref{sec:languages+use}, we already know some things about this type space. When
it comes to communicative success: types with $\Lbound$ have a functional advantage over
$\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable
advantage over literal $\Llack$. The red arrows in Fig.~\ref{fig:sketch} sketch out these
relationships. They depict the rough trajectories we can expect if there is only pressure for
communicative success. How much each replicator step moves a population through this space
ultimately depends on the rationality parameter $\lambda$. In particular, note that functional
differences decrease as $\lambda$ increases: if $\lambda$ is sufficiently high, all types
except for literal $\Llack$ speakers will show a strong tendency to associate $\ssome$ with
$\msome$ and $\sall$ with $\mall$.

As a consequence, when it comes to learnability and transmission fidelity, the likelihood that
any of these three types produces a given datum $d$ will be approximately equal. We can
therefore expect that, if the prior is flat, no type is passed on very faithfully in this
space. This would lead populations to gravitate toward the space's center after mutator
steps. This is sketched by the blue dash-dotted arrows in Fig.~\ref{fig:sketch}. Neither prior
nor likelihood alone, but their combination, are what determines the effects of (iterated)
Bayesian learning. If, instead of a flat prior, there is a prior that favors, say, $\Llack$
over $\Lbound$ we can expect mutator steps to be skewed toward the upper-half of the
space. This is suggested by the solid purple arrows in Fig.~\ref{fig:sketch}. Taking stock, the
effects of iterated learning will depend not only on $\lambda$, affecting the likelihood, but
also on the prior (how strong of a bias there is) and the degree to which learners tend to
maximize the posterior (controlled by parameter $l$).

@@

Fig.~\ref{fig:sketch} and \ref{fig:quiver} depict replicator and mutator steps in isolation to
better showcase their effects. Their combination will yield different outcomes, which we now
turn to in a case study with a larger type space and more motivation for particular assumptions
about learning biases.

@@

Consider a huge, virtually infinite population of agents. At any point in time each agent has a
fixed type $i$ which determines its fitness $f_i$ at that moment in time when playing with an
arbitrary member of the population. The relative frequency of type $i$ is $x_i$ with
$\sum_i x_i = 1$. The average population fitness is $\Phi = \sum_k x_k f_k$. The mutation
probability from $j$ to $i$ is $Q_{ji}$, with $\sum_k Q_{jk} = 1$. The proportion of type $i$
at the next time step $x'_i$ is given by the discrete time replicator dynamic:
\begin{align}
  \label{eq:RMD_discrete_repeated}
  x'_i = \sum_j \frac{x_jf_j Q_{ji}}{\Phi}\,,
\end{align}
We will derive this dynamic from an agent-level update scheme of conditional imitation. There
are many variants on this derivation. Our choice is to be as simple as
possible. \citet{Sandholm2010:Population-Game} provides much more detail and discusses many
interesting variants.

A simple conditional imitation dynamic runs as follows. Every now and then an random agent is
selected at random from the population and given a chance to possibly change its behavior. Call
this agent (perhaps slightly misleadingly) \emph{learner} $L$. The learner chooses a random
agent from the population to possibly imitate. Call this agent (somewhat more misleadingly)
\emph{teacher} $T$. Whether $L$ actually chooses to imitate the behavior of $T$ depends on how
good $T$'s behavior is at the moment. For simplicity, we simply assume that $L$ adopts $T$'s
type with a probability given by $f_j$, where $j$ is the type of $T$.\footnote{We assume here
  that $f_i \in [0;1]$ for all $i$. This is actually the case for the definition of fitness as
  communicative success from Section~\ref{sec:fitn--learn}. See
  \citet{Sandholm2010:Population-Game} for more complex alternative conditional update rules
  that still derive the same population dynamic.} 
If $L$ chooses to imitate $T$'s behavior, it may not be able to directly observe $T$'s actual
type. Rather it may have to infer $T$'s latent type from its overt behavior. The probability
that $L$ then actually acquires type $k$ when it chose to imitate a teacher of type $j$ is
$Q_{jk}$. In sum, we assume that learners have (at least roughly) an estimate of how good
another agent's behavior is, decide to imitate that behavior conditionally on how good it
actually is and the adopt whatever mechanism (type) they infer may cause the observable
behavior.

>>>>

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses, at least since \citeposs{zipf:1949} explanation of
word frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another, as natural languages need to be learnable to survive their
faithful transmission across generations. Furthermore, the effects of even small learning
biases can have striking effects on an evolving language in a process of iterated learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

@@

\subsection{The replicator mutator dynamic}
An elegant formal approach to capture the interaction between selection gor communicative
efficiency and learnability is the \emph{replicator mutator dynamic}
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. The
RMD is a population dynamic. It describes the average change in the composition of a community
of agents of different types. The dynamic assumes a virtually infinite population, so that it
is able to abstract away from what individual agents do. The result is an abstract description
of the most probable trajectory of change in the population vector $\vec{x}$, which codifies
relative frequencies of types, so that $x_i$ is the relative frequency of type $i$ and
$\sum\nolimits_i x_i = 1$. In its simplest discrete-time formulation, the RMD defines the
frequency $x'_i$ of each type $i$ in the population at the next time step as a function of: (i)
the frequency $x_i$ of each type $i$ before the update step, (ii) the fitness $f_i$ of each
type $i$ before the update, and (iii) the probability $Q_{ji}$ that an agent who observes the
behavior of an agent with type $j$ ends up acquiring type $i$:
\begin{align}
  \label{eq:RMD_discrete}
  x'_i = \sum_j \frac{x_jf_j Q_{ji}}{\Phi}\,,
\end{align}
where $\Phi = \sum_k x_k f_k$ is the average fitness in the population.

The RMD consists of two components: fitness-based selection and transmission biases encoded in
the so-called mutation matrix $Q$. This becomes most transparent when we consider an equivalent
formulation in terms of a step-wise application of the discrete-time replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$ and a
subsequent multiplication with the mutation matrix $Q$:
\begin{align}
  \label{eq:RMD_discrete_recast}
  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
\end{align}
where
\begin{align*}
      \left ( \text{RD}(\vec{x}) \right )_i 
         = \frac{x_i f_i}{\Phi}
 \ \ \ \ \text{and} \ \ \ \ 
  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
          x_j Q_{ji} \right)_i\,.
\end{align*}
%
If there is no mutation, so that types are always replicated faithfully with $Q_{ji}=1$
whenever $j=i$, the RMD reduces to the replicator dynamic. If fitness is the same for all
types, $f_i = f_j$ for all $i$ and $j$, the RMD reduces to a process of iterated
mutation. Appendix~\ref{sec:minim-example-repl} provides an abstract minimal example of how the
replicator mutator dynamic works.

@@

According to the replicator dynamic the relative frequency $x_i$ of type $i$ increases
proportional to its average fitness in the population. This dynamic is popular and versatile
because it can be derived from many processes of biological and cultural transmission and
selection \citep[for overview and several derivations
see][]{Sandholm2010:Population-Game}. Under a biological interpretation, the model describes
population-level changes when the number of offspring under single-parent preproduction is
proportional to the parent's fitness. Under a cultural interpretation, examples of concrete
agent-level processes that lead to population-level changes as described by the replicator
dynamic include conditional imitation
\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. In
conditional imitation, for instance, each agent carries a fixed type $i$ which it may change to
a type $j$ when it happens to observe another agent using type $j$. Imitation occurs with a
probability that is a monotonic function of how well $j$ fares against the population: the
better the behavior $j$, the more likely it is to be imitated.\footnote{A crucial assumption in
  all of the derivations of the replicator dynamic as a biological or cultural process of
  evolution is that agents themselves are rather limited in their strategic capacities. They
  are, for example, not able to innovate a strategy, i.e., to just choose a type, that is
  optimal given the current population state. They are also not able to strategically adapt
  their behavior to any particular agent they are interacting with on some given occasion. Such
  minimalism about the strategic capacities of agents is a conservative methodological
  choice. It allows us to study ---as the formal thought experiment that any such model is---
  how fitness-based selection changes a population in the absence of optimal agent-level
  reasoning and it abstracts away from specific assumptions about how fitness-based selection
  might be realized concretely.}

@@

In the following application to language evolution, we prefer a cultural interpretation of the
mutator component as the perturbation induced by how easy different linguistic traits are
learnable from observation of language use. Nothing of current relevance hinges on whether the
replicator component is interpreted as a biological process or a cultural process. We believe
that this flexibility of interpretation is an asset. It abstracts away from specific
assumptions and focuses on a general and versatile description of fitness-based
selection. Still, to have something concrete in mind when interpreting the following
application, Appendix~\ref{sec:deriv-repl-mutat} derives the RMD from a specific scheme of
agent-level conditional imitation.

@@

n the context of language evolution, fitness is usually associated with the ability to
successfully communicate with other language users from the same population
\citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Concretely, the fitness
of type $i$ is its average expected communicative success, or \emph{expected utility} (EU),
given the relative frequencies $x_j$ of types $j$ in the current population:
\begin{align*}
  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
\end{align*}
In signaling games, the expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
average success of $i$ when talking or listening to $j$. If, as standardly assumed, agents are speakers half of the
time, this yields:
\begin{align*}
  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, \text{EU}_S(t_i,t_j) + \nicefrac{1}{2} \, \text{EU}_H(t_i,t_j)\,,
\end{align*}
where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
speaker and as a hearer when communicating with $j$, defined as follows, where $n_i$ and $n_j$
are type $i$'s and type $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
\begin{flalign*}
  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} H_{n_j}(s' \mid m;L_j)
  \delta(s,s') \,, \\
 & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)\,.
\end{flalign*}
As usual, we assume that agents are cooperative: $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.

@@

Communicative success and learnability are central to the cultural evolution of language. These
components can be modelled, respectively, as replication based on a measure of fitness in terms
of communicative efficiency and iterated Bayesian learning. Their interaction is described by
the discrete time replicator mutator dynamic in (\ref{eq:RMD_discrete}), repeated here:
\begin{align*}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_h x_h f_h}\,.
\end{align*}
This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
frequency $x_i$ before the step, its fitness $f_i$, and the probability that a learner infers
$i$ when observing the behavior of a type-$j$ agent. Fitness-based can be thought of as
biological (fitness as expected relative number of offspring) or cultural evolution (fitness as
likelihood of being imitated or repeated). A type's communicative success depends on how well
it communicates within a population while its learnability depends on the fidelity by which it
is inferred by new generations of learners. The learners' task is consequently to perform a
joint inference over types of linguistic behavior and lexical meaning.

@@

This section engages in a formal thought experiment to address the question: if a population of
language users could freely combine different lexica with different dispositions to act on
them, what are conditions under which the majority view of scalar implicatures could have
evolved?

Recall that the majority view is that scalar implicatures are non-lexicalized pragmatic
enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may} are
semantically weak expressions for which logically stronger expressions are salient, e.g.,
\emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em
  all}. Still, the use of a less informative expression (\emph{some}) when a more informative
one (\emph{all}) could have been used can license a defeasible inference (\emph{some but not
  all}) that stronger alternative is false (cf. \citealt{horn:1972,gazdar:1979}). According to
the majority view, this is a pragmatic inference, not part of the conventional meaning.

In the following we consider a specific application of the model from Section~\ref{sec:model}
which allows us to address the question if or when scalar inferences might (not) lexicalize. We
consider what are perhaps two of the simplest non-trivial setups that speak to this matter and
reflect on their limitations in Section~\ref{sec:discussion}. Section \ref{subsec:example}
first considers a minimal example intended to isolate effects of replication and mutation and
to gain insights into the sense in which semantic meaning and pragmatic use can be said to
co-evolve. We then turn to a larger and more realistic setup in
Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes simulations
for this case and their results.

\subsection{Co-evolution of semantics \& pragmatics in a restricted type space}
\label{subsec:example}

@@

@@

From Section~\ref{sec:languages+use}, we already know some things about this type space. When
it comes to communicative success, types with $\Lbound$ have a functional advantage over
$\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable
advantage over literal $\Llack$. The red arrows in Fig.~\ref{fig:sketch} sketch out these
relationships. They depict the rough trajectories we can expect if there is only pressure for
communicative success. How much each replicator step moves a population through this space
ultimately depends on the rationality parameter $\lambda$. Functional differences decrease as
$\lambda$ increases: if $\lambda$ is sufficiently high, all types except for literal $\Llack$
speakers will show a strong tendency to associate $\ssome$ with $\msome$ and $\sall$ with
$\mall$.

As a consequence, when it comes to learnability and transmission fidelity, the likelihood that
any of these three types produces a given datum $d$ will be approximately equal. We can
therefore expect that, if the prior is flat, no type is passed on very faithfully in this
space. This would lead populations to gravitate toward the space's center after mutator
steps. This is sketched by the blue dash-dotted arrows in Fig.~\ref{fig:sketch}. Neither prior
nor likelihood alone, but their combination, are what determines the effects of (iterated)
Bayesian learning. If, instead of a flat prior, there is a prior that favors, say, $\Llack$
over $\Lbound$ we can expect mutator steps to be skewed toward the upper-half of the
space. This is suggested by the solid purple arrows in Fig.~\ref{fig:sketch}. \mf{Should we
  include here a sentence on why the top-most arrow for biased learning points to pragmatic
  L-lack?} Taking stock, the effects of iterated learning will depend not only on $\lambda$,
affecting the likelihood, but also on the prior (how strong of a bias there is) and the degree
to which learners tend to maximize the posterior (controlled by parameter $l$).

@@

Fig.~\ref{fig:sketch} and \ref{fig:quiver} depict replicator and mutator steps in isolation to
better showcase their effects. How exactly these forces shape a population over time also
hinges on which types are present. Especially the effects of learning biases affected by
this. If, hypothetically, we add many functionally dispreferred types which all happen to
mutate with a high probability to the same type, this can strongly bias the dynamics in favor
of this type. \mf{maybe a sentence here about the top-most arrow in Figure 1?} For this reason,
we turn to a larger setup next with a larger and more realistic set of types and a suggestion
for how to formalize potential learning biases in lexical inferences.

Despite its simplicity, the diagram in Figure~\ref{fig:sketch} makes clear the sense in which
semantics and pragmatics might co-evolve. Pragmatic language use will not evolve under a strong
and fully expressive semantics (bottom part of Figure~\ref{fig:sketch}). A weak semantic
convention will not evolve without disposition towards pragmatic enrichments (left part of
Figure~\ref{fig:sketch}). But together a weak semantics and a pragmatic disposition for
enrichment can co-evolve.

@@

Consider a huge, virtually infinite population of agents. At any point in time each agent has a
fixed type $i$ which determines its fitness $f_i$ at that moment in time when playing with an
arbitrary member of the population. The relative frequency of type $i$ is $x_i$ with
$\sum_i x_i = 1$. The average population fitness is $\Phi = \sum_k x_k f_k$. The mutation
probability from $j$ to $i$ is $Q_{ji}$, with $\sum_k Q_{jk} = 1$. The proportion of type $i$
at the next time step $x'_i$ is given by the discrete-time replicator mutator dynamic:
\begin{align}
  \label{eq:RMD_discrete_repeated}
  x'_i = \sum_j \frac{x_jf_j Q_{ji}}{\Phi}\,,
\end{align}
We will derive this dynamic from an agent-level update scheme of conditional imitation. There
are many variants on this derivation. Our choice is to be as simple as
possible. \citet{Sandholm2010:Population-Game} provides much more detail and discusses many
further interesting derivations of similar evolutionary dynamics.

A simple conditional imitation dynamic runs as follows. Every now and then an agent is selected
at random from the population and given a chance to possibly change its behavior. Call this
agent (perhaps slightly misleadingly) \emph{learner} $L$. The learner chooses a random agent
from the population to possibly imitate. Call this agent (somewhat more misleadingly)
\emph{teacher} $T$. Whether $L$ actually chooses to imitate the behavior of $T$ depends on how
good $T$'s behavior is at the moment. For simplicity, we assume that $L$ adopts $T$'s type with
probability $f_j$, where $j$ is the type of $T$.\footnote{We assume here that $f_i \in [0;1]$
  for all $i$. This is actually the case for the definition of fitness as communicative success
  from Section~\ref{sec:fitn--learn}. See \citet{Sandholm2010:Population-Game} for more complex
  alternative conditional update rules that still derive the same population dynamic.}  If $L$
chooses to imitate $T$'s behavior, it may not be able to directly observe $T$'s actual
type. Rather it may have to infer $T$'s latent type from its overt behavior. The probability
that $L$ actually acquires type $k$ when it chose to imitate a teacher of type $j$ is
$Q_{jk}$. In sum, we assume that learners have (at least roughly) an estimate of how good
another agent's behavior is, decide to imitate that behavior conditionally on how good it is
and the adopt whatever mechanism (type) they infer may cause the observed behavior.
